GADTs, Functoriality, Parametricity: Pick Two

### 2021-06-08 20-gadts/21-LSFA/GADT-story.tex
final LSFA paper

Since the shape of an ADT structure --- i.e., a structure whose type
is an instance of an ADT --- is independent of the type of data it
contains, ADTs can be defined polymorphically. As a result, an ADT
structure containing data of type $\mathsf{a}$ can be transformed into
another ADT structure of the exact same shape containing data of
another type $\mathsf{b}$ simply by applying a given function
$\mathsf{f \,::\,a \to b}$ to each of its elements. Moreover, every ADT
$\mathsf{T}$ can be made an instance of Haskell's $\mathsf{Functor}$
class by defining a type-and-data-uniform, structure-preserving,
data-changing function $\mathsf{map_T}$ for it.\footnote{We write
  $\mathsf{map_T}$, or simply $\mathsf{map}$ when $\mathsf{T}$ is
  clear from context, for the function $\mathsf{fmap :: (a \to b) \to
    (T\,a \to T\,b)}$ witnessing that a type constructor $\mathsf{T}$
  is an instance of Haskell's $\mathsf{Functor}$ class. We
  emphasize that $\mathsf{fmap}$ functions in Haskell are intended to
  satisfy syntactic reflections of the functor laws ---
%  iality properties ---
  i.e., preservation of identity functions and composition of
  functions --- even though this is not enforced by the compiler and
  is instead left to the good intentions of the programmer.}
Then, given a type-independent way of rearranging an ADT
structure's shape $\mathsf{T\,a}$ into the shape for another ADT
structure $\mathsf{T'\,a}$, we get the same structure of type
$\mathsf{T'\,b}$ regardless of whether we first rearrange the original
structure of type $\mathsf{T\,a}$ into one of type $\mathsf{T'\,a}$
and then use $\mathsf{map_{T'}}$ to convert that resulting structure
to one of type $\mathsf{T'\,b}$, or we first use $\mathsf{map_T}$ to
convert the original structure of type $\mathsf{T\,a}$ to one of type
$\mathsf{T\,b}$ and then rearrange that resulting structure into one
of type $\mathsf{T'\,b}$. For example, if $\mathsf{f :: a \to b}$,
$\mathsf{t \,::\,Tree\,a}$, and $\mathsf{g :: Tree\,a\to List\,a}$
arranges lists into trees in a type-independent way, then we have the
following rearrange-transform property:
\[\mathsf{map_{List}\,f\,(g \, t) \,=\, g\,(map_{Tree}\,f\,t)}\]

@@

Requiring $F$ to be a functor ensures that the interpretation $\mu F$
of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure that
$\mu F$ is itself a functor, so that the type constructor $\mathsf{T}$
associated with $\mathsf{T\,a}$ also supports its own $\mathsf{map}$
function, we can require that $F$ also be the fixpoint $\mu H$ of a
functor $H$. In this case, however, $H$ must be a functor on the
category $\mathit{Set}^\mathit{Set}$ of functors and natural
transformations on $\mathit{Set}$. That is, $H$ must be a {\em
  higher-order} functor on $\mathit{Set}$. Reflecting this requirement
back into syntax gives that $\mathsf{T\,a} = \mathsf{(\mu H)\,a}$ for
the ``type constructor constructor'' $\mathsf{H}$ that supports
suitable\footnote{The $\mathsf{map}$ function for $\mathsf{H}$ is
  intended to satisfy syntactic reflections of the functor laws in
  $\mathit{Set}^\mathit{Set}$ --- i.e., preservation of identity
  natural transformations and composition of natural transformations
  --- and the $\mathsf{map}$ function for $\mathsf{H\,F}$ is intended
  to satisfy syntactic reflections of the functor laws in
  $\mathit{Set}$, even though there is no mechanism in Haskell for
  enforcing this.} $\mathsf{map}$ functions.

@@

To compute the interpretation of $\mathsf{G}$'s fixpoint
representation we first note that the type of $\mathsf{G}$'s solitary
constructor $\mathsf{C :: G\,1}$ is equivalently expressed as
$\mathsf{C :: 1 \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as
$\mathsf{C :: (Lan_{\lambda u. 1}\,\lambda u. 1)\,a \to G\, a}$, where
$\mathsf{\lambda u. 1}$ is the syntactic reflection of the constantly
$1$-valued functor from the category $\mathit{Set}^0$ with a single
object to $\mathit{Set}$. We can therefore represent $\mathsf{G}$ as
\[\mathsf{G\,a} = \mathsf{(\mu \phi. \lambda b.
 (Lan_{\lambda u. 1} \,\lambda u.  1)\,b)\,a}\] The interpretation of
$\mathsf{G}$ is therefore obtained by computing the fixpoint of the
interpretation of the body $\mathsf{\lambda b. (Lan_{\lambda u. 1}
  \,\lambda u. 1)\,b}$ of the syntactic fixpoint $\mathsf{\mu
  \phi. \lambda b.  (Lan_{\lambda u. 1} \,\lambda u.  1)\,b}$ and
applying the result to $\mathsf{a}$. But since the recursion variable
$\mathsf{\phi}$ does not appear in this body, the interpretation of
the fixpoint is just the interpretation of the body itself.  The
interpretation of $\mathsf{G\,a}$ is therefore $(\mathit{Lan}_{\lambda
  u. 1} \lambda u. 1)\, A$, where $A$ interprets $\mathsf{a}$.
%and $1$ is the {\color{blue} constantly $1$-valued
%  functor $\ast \to \set$ from the category with a single object
%  $\ast$ to $\set$.}
It turns out, however, that, for any set $A$, $(\mathit{Lan}_{\lambda
  u. 1} \lambda u. 1)\,A$ is, in fact, exactly $A$. Indeed,
Proposition~7.1 of~\cite{blw03} gives that $(\mathit{Lan}_{\lambda
  u. 1} \lambda u. 1)\,A$ can be computed as the set of all functions
$a : 1 \to A$ modulo the smallest equivalence relation $\sim$ such
that, for any $f : 1 \to 1$, we have $a \sim a \circ f$.  But this set
is isomorphic to $A$, because the only function $f : 1 \to 1$ is the
identity function. This is different from the interpretation of
$\mathsf{G}$'s Church encoding from the previous paragraph.

Putting it all together, we see that the Church encoding of
$\mathsf{G}$ and its fixpoint representation are not semantically
equivalent: the interpretation of the former has exactly one data
element at instance $\mathsf{G\,1}$ and no elements at any other
instances, whereas the interpretation of the latter has data elements
at every instance other than $\mathsf{G\,0}$.
%Indeed,
These additional data elements can be obtained by reflecting back into
syntax the elements $\mathit{map}_G\,f_a\,c : G\,A$ resulting from
applying the functorial action $\mathit{map}_G$ of $\mathsf{G}$'s
interpretation $G$ to the functions $f_a : 1 \to A$ determined by the
elements $a$ of $A \not = \emptyset$ and the interpretation $c$ of
$\mathsf{C}$.
\end{example}

Forced to choose, a programmer would likely find the idea that a GADT
contains data not specified by its syntax more than a little
disturbing. What, they might ask, should a data type contain other
than data that are constructed using its data constructors? Why should
a GADT contain ``hidden'' elements that are not specified by the
GADT's syntax and are only accessible via applications of
$\mathsf{map}$? From a semanticist's point of view, however, the
prinitive representation of GADTs is entirely reasonable. Indeed, they
would likely find the nonfunctorial nature of a GADT's Church encoding
unnerving at best. After all, they would likely argue, the data in a
GADT shouldn't change or become ill-typed just because a function is
mapped over it. The fact that this happens when GADTs are represented
by their Church encodings actually highlights how GADTs {\em do not}
generalize the essential, container-ish nature of ADTs at all. A
semanticist might therefore conclude that GADTs are seriously
misnamed.
%They might even go on to wonder: in what sense do GADTs
%really deserve to be considered {\em data types} at all?

@@

\section{Parametricity in the Presence of GADTs}\label{sec:par}

{\em Relational parametricity} encodes a powerful notion of type
uniformity, or representation independence, for data types in
functional languages. It formalizes the intuition that a polymorphic
program must act uniformly on all of its possible type instantiations
by requiring that every such program preserves all relations between
pairs of types at which it is instantiated. Parametricity was
originally put forth by Reynolds~\cite{rey83} for System F. It was
later popularized as Wadler's ``theorems for free''~\cite{wad89}, so
called because it can deduce properties of programs solely from their
types, i.e., with no knowledge whatsoever of the text of the programs
involved.  Most of Wadler's free theorems are consequences of
naturality for polymorphic list-processing functions. However,
parametricity can also derive results that go beyond just naturality,
such as inhabitation results, and prove the equivalence of Church
encodings and fixpoint representations of ADTs and
%hereditarily polynomial
nested types by validating short cut fusion and other program
equivalences for them.

To discuss relational parametricity we will need to interpret data
types not just in $\mathit{Set}$, but in a suitable category of
relations as well. The following definition is standard:
\begin{definition}
  The category $\mathit{Rel}$ has:
  \begin{itemize}
\item objects: A relation is a triple $(A,B,R)$, where $R$ is a
  subset of $A \times B$.
\item morphisms: A morphism from $(A,B,R)$ to $(A',B',R')$ is a pair
  $(f : A \to A',g : B \to B')$ of functions in $\mathit{Set}$ such
  that $(f a,g\,b) \in R'$ if $(a,b) \in R$.
\item identities: The identity morphism on $(A,B,R)$ is the pair
  $(\mathit{id}_A : A \to A, \mathit{id}_B : B \to B)$.
\item composition: Composition is the componentwise composition in
  $\mathit{Set}$. That is, $(g_1,g_2) \circ (f_1,f_2) = (g_1 \circ
  f_1, g_2 \circ f_2)$, where the composition being defined on the
  left-hand side is in $\mathit{Rel}$, and the two componentwise
  compositions on the right-hand side are in $\mathit{Set}$.
\end{itemize}
\end{definition}
\noindent
We write $R : \mathit{Rel}\,(A,B)$ for $(A,B,R) \in \mathit{Rel}$.  If
$R : \mathit{Rel}\,(A,B)$ then we write $\pi_1 R$ and $\pi_2 R$ for
the {\em domain} $A$ and {\em codomain} $B$ of $R$, respectively. We
write $\mathit{Eq}_A = (A,A,\{(x,x)~|~ x \in A\})$ for the {\em
  equality relation} on the set $A$.

The key idea underlying parametricity is to give each type
$\mathsf{G[a]}$\footnote{The notation $\mathsf{G[a]}$ indicates that
  $\mathsf{G}$ is a type with one hole which has been filled with the
  type $\mathsf{a}$.}  with one free variable $\mathsf{a}$ a {\em set
  interpretation} $G_0$ taking sets to sets and a \emph{relational
  interpretation} $G_1$ taking relations $R : \mathit{Rel}\,(A,B)$ to
relations $G_1 \,R : \mathit{Rel}\,(G_0 \,A, G_0 \,B)$, and to
interpret each term $\mathsf{t\,(a,x) :: G[a]}$ with one free term
variable $\mathsf{x :: F[a]}$ as a function $t$ associating to each
set $A$ a morphism $t \,A : F_0\,A \to G_0\,A$ in $\mathit{Set}$.
These interpretations are given inductively on the structures of
$\mathsf{G}$ and $\mathsf{t}$ in such a way that they imply two
fundamental theorems. The first is an \emph{Identity Extension Lemma},
which states that $G_1\,\mathit{Eq}_A = \mathit{Eq}_{G_0 A}$, and is
the essential property that makes a model relationally parametric
rather than just induced by a logical relation.  The second is an
\emph{Abstraction Theorem}, which states that, for any $R
:\mathit{Rel}\,(A, B)$, $(t\, A, t\,B)$ is a morphism in
$\mathit{Rel}$ from $(F_0\,A,F_0\,B,F_1\,R)$ to
$(G_0\,A,G_0\,B,G_1\,R)$. The Identity Extension Lemma is similar to
the Abstraction Theorem except that it holds for {\em all} elements of
a type's interpretation, not just those that interpret terms.  Similar
theorems are required for types and terms with any number of free
variables.

The existence of parametric models that interpret GADTs as the
interpretations of their Church encodings follows from, e.g., the
existence of the parametric model of $F_\omega$ constructed
in~\cite{atk12}.  In that model, types are interpreted ``in parallel''
in (types corresponding to) $\mathit{Set}$ and $\mathit{Rel}$ in the
usual way, including the familiar ``cutting down'' of the
interpretations of $\forall$-types to just those elements that are
``parametric''~\cite{rey83,wad89} to ensure that the Identity
Extension Lemma holds. If the set interpretation of $\mathsf{Equal}$
is the function $\mathit{Equal}$ and if the relational interpretation
of a closed type $\mathsf{a}$ with interpretation $A$ is
$\mathit{Eq_A}$ as intended, then the parametricity property for a
GADT is an inhabitation result saying that the set interpreting any
instance of that GADT contains exactly the interpretations of the data
elements that can be formed using its data constructors and whose type
is that instance. In terms of syntax the parametricity property for a
GADT's Church encoding gives that the GADT is inhabited at any
instance for which data elements can be formed using its data
constructors. In particular, the parametricity property for the GADT
$\mathsf{G}$ from~\eqref{eq:G} gives that $\mathsf{G\,a}$ contains a
single data element if $\mathsf{a}$ is semantically equivalent to
$\mathsf{1}$ and none otherwise:

\begin{example}\label{ex:CE-par}
Let $\mathsf{t}$ be a term of type $\mathsf{G\,a}$ for the GADT
$\mathsf{G}$ defined in~\eqref{eq:G}, let $G = (G_0,G_1)$ be the
interpretation of the Church encoding of $\mathsf{G}$, let $t$ be the
interpretation of $\mathsf{t}$, and let $R :
\mathit{Rel}\,(A,B)$. Then $t\, A \in G_0\,A$ and $t\,B \in G_0\,B$
and, by the Abstraction Theorem (Theorem~3) in~\cite{atk12}, $t\,A$
and $t\,B$ must be related in $G_1\,R$. However, under the semantics
given in~\cite{atk12}, which includes the aforementioned
interpretations of $\mathsf{Equal}$ and closed types, the relational
interpretation $G_1\,R$ of $\mathsf{G}$ is itself $\mathit{Eq}_1$ when
$R$ is the relational interpretation $\mathit{Eq}_1$ of $\mathsf{1}$,
and the empty relation whenever $R$ differs from
$\mathit{Eq}_1$. Reflecting back into syntax we deduce that there can
be no term in the type that is the Church encoding of $\mathsf{G\,a}$
unless $\mathsf{a}$ is semantically equivalent to $\mathsf{1}$.
\end{example}

@@

Intuitively, when $\mathsf{G}$ is viewed as a fixpoint, its data
elements include those given by functorial completion. Thus, given any
relation $S : \mathit{Rel}\,(C, D)$ and any morphism $m : S \to R$,
$G_1\, R$ must contain all elements of the form $G_1\,m\,x$ for $x \in
G_1\,S$.  But the two components $m_1 : C \to A$ and $m_2 : D \to B$
of $m$ cannot be given independently of one another, since
%$m$ has to send elements related by $S$ to elements
%related by $R$ in order to be a well-defined morphism of relations.
$(m_1\,c, m_2\,d)$ must be in $R$ whenever $(c,d)$ is in $S$ for
$(m_1,m_2)$ to be a well-defined morphism of relations.  The domain of
$G_1\, R$ thus depends on both $A$ and $B$, rather than simply on $A$.
Likewise, the codomain of $G_1\,R$ also depends on both $A$ and $B$.
The domain and codomain therefore cannot simply be $G_0\, A$ and
$G_0\, B$, respectively.  This suggests that GADTs might fail to have
relational interpretation, and thus might fail to have a parametric
model, as described in the previous paragraph.
%But, as we now show, this is not generally possible when
%GADTs are interpreted as the interpretations of their primitive
%representations.

>>>>

Since the shape of an ADT structure --- i.e., a structure whose type
is an instance of an ADT --- is independent of the type of data it
contains, ADTs can be defined polymorphically. As a result, an ADT
structure containing data of type $\mathsf{a}$ can be transformed into
another ADT structure of the exact same shape containing data of
another type $\mathsf{b}$ simply by applying a given function
$\mathsf{f \,::\,a \to b}$ to each of its elements. Moreover, every ADT
$\mathsf{T}$ can be made an instance of Haskell's $\mathsf{Functor}$
class by defining a type-and-data-uniform, structure-preserving,
data-changing function $\mathsf{map_T}$ for it.\footnote{We write
  $\mathsf{map_T}$, or simply $\mathsf{map}$ when $\mathsf{T}$ is
  clear from context, for the function $\mathsf{fmap :: (a \to b) \to
    (T\,a \to T\,b)}$ witnessing that a type constructor $\mathsf{T}$
  is an instance of Haskell's $\mathsf{Functor}$ class. We
  emphasize that $\mathsf{fmap}$ functions in Haskell are intended to
  satisfy syntactic reflections of the functor laws ---
%  iality properties ---
  i.e., preservation of identity functions and composition of
  functions --- even though this is not enforced by the compiler and
  is instead left to the good intentions of the programmer.}
Then, given a type-independent way of rearranging an ADT
structure's shape $\mathsf{T\,a}$ into the shape for another ADT
structure $\mathsf{T'\,a}$, we get the same structure of type
$\mathsf{T'\,b}$ regardless of whether we first rearrange the original
structure of type $\mathsf{T\,a}$ into one of type $\mathsf{T'\,a}$
and then use $\mathsf{map_{T'}}$ to convert that resulting structure
to one of type $\mathsf{T'\,b}$, or we first use $\mathsf{map_T}$ to
convert the original structure of type $\mathsf{T\,a}$ to one of type
$\mathsf{T\,b}$ and then rearrange that resulting structure into one
of type $\mathsf{T'\,b}$. For example, if $\mathsf{f :: a \to b}$,
$\mathsf{t \,::\,Tree\,a}$, and $\mathsf{g :: Tree\,a\to List\,a}$
arranges trees into lists in a type-independent way, then we have the
following rearrange-transform property:
\[\mathsf{map_{List}\,f\,(g \, t) \,=\, g\,(map_{Tree}\,f\,t)}\]

@@

Requiring $F$ to be a functor ensures that the interpretation $\mu F$
of the ADT $\mathsf{T\,a} = \mathsf{\mu F}$ exists. But to ensure that
$\mu F$ is itself a functor, so that the type constructor $\mathsf{T}$
associated with $\mathsf{T\,a}$ also supports its own $\mathsf{map}$
function, we can require that $F$ be a functor on the category
$\mathit{Set}^\mathit{Set}$ of functors and natural transformations on
$\mathit{Set}$. That is, $F$ must be a {\em higher-order} functor on
$\mathit{Set}$. Writing $H$ in place of $F$ to emphasize that it is
higher-order, and reflecting this requirement back into syntax, we
have that $\mathsf{T\,a} = \mathsf{(\mu H)\,a}$ for the ``type
constructor constructor'' $\mathsf{H}$ that supports
suitable\footnote{The $\mathsf{map}$ function for $\mathsf{H}$ is
  intended to satisfy syntactic reflections of the functor laws in
  $\mathit{Set}^\mathit{Set}$ --- i.e., preservation of identity
  natural transformations and composition of natural transformations
  --- and the $\mathsf{map}$ function for $\mathsf{H\,F}$ is intended
  to satisfy syntactic reflections of the functor laws in
  $\mathit{Set}$, even though there is no mechanism in Haskell for
  enforcing this.} $\mathsf{map}$ functions.

@@

To compute the interpretation of $\mathsf{G}$'s fixpoint
representation we first note that the type of $\mathsf{G}$'s solitary
constructor $\mathsf{C :: G\,1}$ is equivalently expressed as
$\mathsf{C :: 1 \to G\,1}$ or, using~\eqref{eq:nat-transfs}, as
$\mathsf{C :: (Lan_{\lambda u. 1}\,\lambda u. 1)\,a \to G\, a}$, where
$\mathsf{\lambda u. 1}$ is the syntactic reflection of the constantly
$1$-valued functor from the category $\mathit{Set}^0$ with a single
object to $\mathit{Set}$. We can therefore represent $\mathsf{G}$ as
\[\mathsf{G\,a} = \mathsf{(\mu \phi. \lambda b.
 (Lan_{\lambda u. 1} \,\lambda u.  1)\,b)\,a}\] The interpretation of
$\mathsf{G}$ is therefore obtained by computing the fixpoint of the
interpretation of the body $\mathsf{\lambda b. (Lan_{\lambda u. 1}
  \,\lambda u. 1)\,b}$ of the syntactic fixpoint $\mathsf{\mu
  \phi. \lambda b.  (Lan_{\lambda u. 1} \,\lambda u.  1)\,b}$ and
applying the result to $\mathsf{a}$. But since the recursion variable
$\mathsf{\phi}$ does not appear in this body, the interpretation of
the fixpoint is just the interpretation of the body itself.  The
interpretation of $\mathsf{G\,a}$ is therefore $(\mathit{Lan}_{\lambda
  u. 1} \lambda u. 1)\, A$, where $A$ interprets $\mathsf{a}$.
%and $1$ is the {\color{blue} constantly $1$-valued
%  functor $\ast \to \set$ from the category with a single object
%  $\ast$ to $\set$.}
It turns out, however, that, for any set $A$, $(\mathit{Lan}_{\lambda
  u. 1} \lambda u. 1)\,A$ is, in fact, exactly $A$. Indeed,
Proposition~7.1 of~\cite{blw03} gives that $(\mathit{Lan}_{\lambda
  u. 1} \lambda u. 1)\,A$ can be computed as
\[\big( \bigcup_{U : \mathit{Set}^0,\,f: (\lambda u.\,1)\,U \to A} (\lambda
u. 1)\,U \;\big) \,/\,\sim
\;\;\;\;=\;\;\;\;
\big( \bigcup_{U :
  \mathit{Set}^0,\,f: 1 \to A} 1 \big)\,/\,\sim\]
%\;\;\;\;=\;\;\;\;
%\big( \bigcup_{f: (\lambda u.1)\,U \to A} 1 \big)\,/\,\sim\]
where $U$ is the unique object of $\mathit{Set}^0$, $*$ is the unique
element of the singleton set $1$, and $\sim$ is the smallest
equivalence relation such that $(U,f, *)$ and $(U,f',*)$ are related
if
\[
\begin{tikzcd}
(\lambda u.1)\,U \ar[rr, "{\mathit{(\lambda u. 1)\,id_U}}"] \ar[dr, "f"']
&& (\lambda u.1)\,U \ar[dl, "f'"] \\
&A
\end{tikzcd} \;\;\; =\;\;\;
\begin{tikzcd}
1 \ar[rr, "{\mathit{id_1}}"] \ar[dr, "f"']
&&1 \ar[dl, "f'"] \\
&A
\end{tikzcd}
\]
commutes, i.e., if $f = f'$. Since the relation generating $\sim$ is
already an equivalence relation, we have that $(U,f,*) \sim (U,f', *)$
iff $f = f'$. Thus, up to isomorphism, $(\mathit{Lan}_{\lambda u. 1}
\lambda u. 1)\,A = \{ f : 1 \to A\}$, i.e., $(\mathit{Lan}_{\lambda
  u. 1} \lambda u. 1)\,A = A$. This is different from the
interpretation of $\mathsf{G}$'s Church encoding from the previous
paragraph.

Putting it all together, we see that the Church encoding of
$\mathsf{G}$ and its fixpoint representation are not semantically
equivalent: the interpretation of the former has exactly one data
element at instance $\mathsf{G\,1}$ and no elements at any other
instances, whereas the interpretation of the latter has data elements
at every instance other than $\mathsf{G\,0}$.
%Indeed,
These additional data elements can be obtained by reflecting back into
syntax the elements $\mathit{map}_G\,f_a\,c \in G\,A$ resulting from
applying the functorial action $\mathit{map}_G$ of $\mathsf{G}$'s
interpretation $G$ to the functions $f_a : 1 \to A$ determined by the
elements $a$ of $A \not = \emptyset$ and the interpretation $c$ of
$\mathsf{C}$.
\end{example}

Forced to choose, a programmer would likely find the idea that a GADT
contains data not specified by its syntax more than a little
disturbing. What, they might ask, should a data type contain other
than data that are constructed using its data constructors? Why should
a GADT contain ``hidden'' elements that are not specified by the
GADT's syntax and are only accessible via applications of
$\mathsf{map}$? From a semanticist's point of view, however, the
primitive representation of GADTs is entirely reasonable. Indeed, they
would likely find the nonfunctorial nature of a GADT's Church encoding
unnerving at best. After all, they would likely argue, the data in a
GADT shouldn't change or become ill-typed just because a function is
mapped over it. The fact that this happens when GADTs are represented
by their Church encodings actually highlights how GADTs {\em do not}
generalize the essential, container-ish nature of ADTs at all. A
semanticist might therefore conclude that GADTs are seriously
misnamed.
%They might even go on to wonder: in what sense do GADTs
%really deserve to be considered {\em data types} at all?

@@

\section{Parametricity in the Presence of GADTs}\label{sec:par}
{\em Relational parametricity} encodes a powerful notion of type
uniformity, or representation independence, for data types in
functional languages. It formalizes the intuition that a polymorphic
program must act uniformly on all of its possible type instantiations
by requiring that every such program preserves all relations between
pairs of types at which it is instantiated. Parametricity was
originally put forth by Reynolds~\cite{rey83} for System F. It was
later popularized as Wadler's ``theorems for free''~\cite{wad89}, so
called because it can deduce properties of programs solely from their
types, i.e., with no knowledge whatsoever of the text of the programs
involved.  Most of Wadler's free theorems are consequences of
naturality for polymorphic list-processing functions. However,
parametricity can also derive results that go beyond just naturality,
such as inhabitation results, and prove the equivalence of Church
encodings and fixpoint representations of ADTs and
%hereditarily polynomial
nested types by validating short cut fusion and other program
equivalences for them.

To discuss relational parametricity we will need to interpret data
types not just in $\mathit{Set}$, but in a suitable category of
relations as well. The following definition is standard:
\begin{definition}\label{def:rel}
  The category $\mathit{Rel}$ has:
  \begin{itemize}
\item objects: A relation is a triple $(A,B,R)$, where $R$ is a
  subset of $A \times B$.
\item morphisms: A morphism from $(A,B,R)$ to $(A',B',R')$ is a pair
  $(f : A \to A',g : B \to B')$ of functions in $\mathit{Set}$ such
  that $(f a,g\,b) \in R'$ if $(a,b) \in R$.
\item identities: The identity morphism on $(A,B,R)$ is the pair
  $(\mathit{id}_A : A \to A, \mathit{id}_B : B \to B)$.
\item composition: Composition is the componentwise composition in
  $\mathit{Set}$. That is, $(g_1,g_2) \circ (f_1,f_2) = (g_1 \circ
  f_1, g_2 \circ f_2)$, where the composition being defined on the
  left-hand side is in $\mathit{Rel}$, and the two componentwise
  compositions on the right-hand side are in $\mathit{Set}$.
\end{itemize}
\end{definition}
\noindent
We write $R : \mathit{Rel}\,(A,B)$ for $(A,B,R) \in \mathit{Rel}$.  If
$R : \mathit{Rel}\,(A,B)$ then we write $\pi_1 R$ and $\pi_2 R$ for
the {\em domain} $A$ and {\em codomain} $B$ of $R$, respectively. We
write $\mathit{Eq}_A = (A,A,\{(x,x)~|~ x \in A\})$ for the {\em
  equality relation} on the set $A$.

The key idea underlying parametricity is to give each type
$\mathsf{G[a]}$\footnote{The notation $\mathsf{G[a]}$ indicates that
  $\mathsf{G}$ is a type with one hole which has been filled with the
  type $\mathsf{a}$.}  with one free variable $\mathsf{a}$ a {\em set
  interpretation} $G_0$ taking sets to sets and a \emph{relational
  interpretation} $G_1$ taking relations $R : \mathit{Rel}\,(A,B)$ to
relations $G_1 \,R : \mathit{Rel}\,(G_0 \,A, G_0 \,B)$, and to
interpret each term $\mathsf{t\,(a,x) :: G[a]}$ with one free term
variable $\mathsf{x :: F[a]}$ as a function $t$ associating to each
set $A$ a morphism $t \,A : F_0\,A \to G_0\,A$ in
$\mathit{Set}$. Here, $F_0$ is the set interpretation of $\mathsf{F}$.
These interpretations are given inductively on the structures of
$\mathsf{G}$ and $\mathsf{t}$ in such a way that they imply two
fundamental theorems. The first is an \emph{Identity Extension Lemma},
which states that $G_1\,\mathit{Eq}_A = \mathit{Eq}_{G_0 A}$, and is
the essential property that makes a model relationally parametric
rather than just induced by a logical relation.  The second is an
\emph{Abstraction Theorem}, which states that, for any $R
:\mathit{Rel}\,(A, B)$, $(t\, A, t\,B)$ is a morphism in
$\mathit{Rel}$ from $(F_0\,A,F_0\,B,F_1\,R)$ to
$(G_0\,A,G_0\,B,G_1\,R)$. The Identity Extension Lemma is similar to
the Abstraction Theorem except that it holds for {\em all} elements of
a type's interpretation, not just those that interpret terms.  Similar
theorems are required for types and terms with any number of free
variables. In particular, if $\mathsf{t}$ is closed (i.e., has no free
term variables) then $t\,A \in G_0\,A$.

The existence of parametric models that interpret GADTs as the
interpretations of their Church encodings follows from, e.g., the
existence of the parametric model of $F_\omega$ constructed
in~\cite{atk12}.  In that model, types are interpreted ``in parallel''
in (types corresponding to) $\mathit{Set}$ and $\mathit{Rel}$ in the
usual way, including the familiar ``cutting down'' of the
interpretations of $\forall$-types to just those elements that are
``parametric''~\cite{rey83,wad89} to ensure that the Identity
Extension Lemma holds. If the set interpretation of $\mathsf{Equal}$
is the function $\mathit{Equal}$ and if the relational interpretation
of a closed type $\mathsf{a}$ with interpretation $A$ is
$\mathit{Eq_A}$ as intended, then the parametricity property for a
GADT is an inhabitation result saying that the set interpreting any
instance of that GADT contains exactly the interpretations of the data
elements that can be formed using its data constructors and whose type
is that instance. The parametricity property for the Church encoding
of a GADT $\mathsf{G}$ gives that $\mathsf{G\,a}$ is inhabited iff data
elements of the instance of $\mathsf{G}$ at $\mathsf{a}$ can be formed
using $\mathsf{G}$'s data constructors.  In particular, the
parametricity property for the GADT $\mathsf{G}$ from~\eqref{eq:G}
gives that $\mathsf{G\,a}$ contains a single data element if
$\mathsf{a}$ is semantically equivalent to $\mathsf{1}$ and none
otherwise:

\begin{example}\label{ex:CE-par}
Let $\mathsf{t}$ be a closed term of type $\mathsf{G\,a}$ for the GADT
$\mathsf{G}$ defined in~\eqref{eq:G}, let $G = (G_0,G_1)$ be the
interpretation of the Church encoding of $\mathsf{G}$, let $t$ be the
interpretation of $\mathsf{t}$, and let $R :
\mathit{Rel}\,(A,B)$. Then $t\, A \in G_0\,A$ and $t\,B \in G_0\,B$
and, by the Abstraction Theorem (Theorem~3) in~\cite{atk12}, $t\,A$
and $t\,B$ must be related in $G_1\,R$. However, under the semantics
given in~\cite{atk12}, which includes the aforementioned
interpretations of $\mathsf{Equal}$ and closed types, the relational
interpretation $G_1\,R$ of $\mathsf{G}$ is itself $\mathit{Eq}_1$ when
$R$ is the relational interpretation $\mathit{Eq}_1$ of $\mathsf{1}$,
and the empty relation whenever $R$ differs from
$\mathit{Eq}_1$. Reflecting back into syntax we deduce that there can
be no term in the type that is the Church encoding of $\mathsf{G\,a}$
unless $\mathsf{a}$ is semantically equivalent to $\mathsf{1}$.
\end{example}

@@

Intuitively, when $\mathsf{G}$ is viewed as a fixpoint, its data
elements include those given by functorial completion. Since $G_1$ is
a functor, given any relation $S : \mathit{Rel}\,(C, D)$ and any
morphism $m : S \to R$, $G_1\, R$ must contain all elements of the
form $G_1\,m\,x$ for $x \in G_1\,S$.  But the two components $m_1 : C
\to A$ and $m_2 : D \to B$ of $m$ cannot be given independently of one
another, since Definition~\ref{def:rel} entails that
%$m$ has to send elements related by $S$ to elements
%related by $R$ in order to be a well-defined morphism of relations.
$(m_1\,c, m_2\,d)$ must be in $R$ whenever $(c,d)$ is in $S$ for
$(m_1,m_2)$ to be a well-defined morphism of relations.  The domain of
$G_1\, R$ thus depends on both $A$ and $B$, rather than simply on $A$.
Likewise, the codomain of $G_1\,R$ also depends on both $A$ and $B$.
The domain and codomain therefore cannot simply be $G_0\, A$ and
$G_0\, B$, respectively.  This suggests that GADTs might fail to have
relational interpretation, and thus might fail to have a parametric
model, as described in the previous paragraph.
%But, as we now show, this is not generally possible when
%GADTs are interpreted as the interpretations of their primitive
%representations.
